---
title: "Air temperature heatwaves"
author: "Sam Bashevkin"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float:
      collapsed: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(tigris_use_cache = TRUE)
```

```{r, message=FALSE}
library(dplyr)
library(tidyr)
library(readr)
library(lubridate)
library(ggplot2)
library(zipcodeR)
library(DT)
library(patchwork)
library(tigris)
library(sf)
```

# Load data

Find zip codes of interest
```{r}
zips<-bind_rows(
  search_city('Stockton','CA'),
  search_city('Sacramento','CA'),
  search_city('West Sacramento','CA'),
  search_city('Pittsburg','CA'),
  search_city('Isleton','CA'))%>%
  select(zipcode, major_city, population)
```

```{r}
zip_gps<-read_csv("shapefiles/zip_gps.csv")%>%
  mutate(ZCTA5=as.character(ZCTA5))%>%
  left_join(zips, by=c("ZCTA5"="zipcode"))%>%
  filter(!is.na(major_city))%>%
  st_as_sf(coords=c("lon", "lat"), crs=4326)%>%
  group_by(major_city)%>%
  summarise()%>%
  st_convex_hull()

ggplot()+
  geom_sf(data=zip_gps, aes(fill=major_city), alpha=0.4)+
  theme_bw()
```


Load humidity data
```{r}
hum_raw<-read_csv("data_clean/gridmet_humidity_CAzip_99to22.csv", 
                  col_types=cols_only(zip="c", date="D",
                                      lat="d", lon="d", rmax="d", 
                                      rmin="d", sph="d"))

hum<-hum_raw%>%
  filter(zip%in%zips$zipcode)%>%
  left_join(zips, by=c("zip"="zipcode"))%>%
  mutate(rh=(rmax+rmin)/2)%>% # Calculate rh with midpoint of min and max
  group_by(date, major_city)%>%
  summarise(rh=mean(rh), .groups="drop")

```


Load temperature data
```{r}
temp_max_raw<-read_csv("data_clean/gridmet_temperature_CAzip_99to22.csv", 
                       col_types=cols_only(zip="c", date="D",
                                           lat="d", lon="d", tmmx_C="d"))

temp_appt<-temp_max_raw%>%
  filter(zip%in%zips$zipcode)%>%
  left_join(zips, by=c("zip"="zipcode"))%>%
  group_by(date, major_city)%>%
  summarise(temp_max=max(tmmx_C), .groups="drop")%>%
  left_join(hum, temp_max, by=c("date", "major_city"))%>%
  mutate(satvap=6.112*10^(7.5*temp_max/(237.7+temp_max)),
         actvap=(rh*satvap)/100,
         dewpt=(-430.22+237.7*log(actvap))/(-log(actvap)+19.08),
         temp_appt=-2.653+(0.994*temp_max)+0.0153*(dewpt*dewpt),
         major_city=factor(major_city, levels=c("Pittsburg", "Isleton", "West Sacramento", "Sacramento", "Stockton")),
         Month=month(date),
         Year=year(date))%>%
  group_by(Month, major_city)%>%
  mutate(temp_appt_q99=quantile(temp_appt, 0.99),
         temp_appt_q95=quantile(temp_appt, 0.95),
         temp_appt_q90=quantile(temp_appt, 0.90))%>%
  ungroup()%>%
  group_by(major_city)%>%
  mutate(temp_appt_l1=lag(temp_appt, 1, order_by=date),
         temp_appt_l2=lag(temp_appt, 2, order_by=date),
         temp_appt_ln1=lead(temp_appt, 1, order_by=date),
         temp_appt_ln2=lead(temp_appt, 2, order_by=date),
         heatwave_99=if_else((temp_appt>temp_appt_q99 & temp_appt_l1>temp_appt_q99 & temp_appt_l2>temp_appt_q99) | # Current day and prior 2 days above cutoff
                               (temp_appt>temp_appt_q99 & temp_appt_ln1>temp_appt_q99 & temp_appt_ln2>temp_appt_q99) | # Current day and next 2 days above cutoff
                               (temp_appt>temp_appt_q99 & temp_appt_ln1>temp_appt_q99 & temp_appt_l1>temp_appt_q99),  # current day and day before and after are above cutoff
                             TRUE, FALSE),
         heatwave_95=if_else((temp_appt>temp_appt_q95 & temp_appt_l1>temp_appt_q95 & temp_appt_l2>temp_appt_q95) |
                               (temp_appt>temp_appt_q95 & temp_appt_ln1>temp_appt_q95 & temp_appt_ln2>temp_appt_q95) |
                               (temp_appt>temp_appt_q95 & temp_appt_ln1>temp_appt_q95 & temp_appt_l1>temp_appt_q95), 
                             TRUE, FALSE),
         heatwave_90=if_else((temp_appt>temp_appt_q90 & temp_appt_l1>temp_appt_q90 & temp_appt_l2>temp_appt_q90) |
                               (temp_appt>temp_appt_q90 & temp_appt_ln1>temp_appt_q90 & temp_appt_ln2>temp_appt_q90) |
                               (temp_appt>temp_appt_q90 & temp_appt_ln1>temp_appt_q90 & temp_appt_l1>temp_appt_q90), 
                             TRUE, FALSE),
         heatwave_95_intensity=if_else(heatwave_95, temp_appt/temp_appt_q95, NA))%>%
  ungroup()%>%
  mutate(Month=month(Month, label=T))%>%
  arrange(major_city, date)

# identifying contiguous 95% heatwave days to estimate duration
heatwave_dur<-temp_appt%>%
  filter(heatwave_95)%>%
  group_by(major_city)%>%
  mutate(Lag=date-lag(date, order_by = date))%>%
  ungroup()%>%
  mutate(Start=if_else(is.na(Lag) | Lag>1, TRUE, FALSE),
         Heatwave_id=1:n(),
         Heatwave_id=if_else(Start, Heatwave_id, NA_integer_),
         Heatwave_id=as.integer(as.factor(Heatwave_id)))%>%
  fill(Heatwave_id, .direction="down")%>%
  group_by(major_city, Heatwave_id)%>%
  summarise(duration=n(),
            Date=min(date),
            temp_appt_mean=mean(temp_appt),
            intensity_mean=mean(heatwave_95_intensity),
            .groups="drop")%>%
  mutate(Month=month(Date, label=T),
         Year=year(Date),
         duration=if_else(duration<3, 3, duration))

heatwave_dur_year<-heatwave_dur%>%
  group_by(Year, major_city)%>%
  summarise(duration_mean=mean(duration),
            duration_sd=sd(duration),
            .groups="drop")%>%
  mutate(duration_sd=if_else(is.na(duration_sd), 0, duration_sd))

heatwave_sum_month<-temp_appt%>%
  group_by(major_city, Month)%>%
  summarise(n = n(),
            freq_99=sum(heatwave_99)/n(),
            freq_95=sum(heatwave_95)/n(),
            freq_90=sum(heatwave_90)/n(),
            .groups="drop")%>%
  pivot_longer(cols = c(freq_99, freq_95, freq_90), values_to="heatwave_freq", names_to="quantile", names_prefix="freq_")%>%
  mutate(quantile=as.integer(quantile))

heatwave_sum_year<-temp_appt%>%
  group_by(major_city, Year)%>%
  summarise(n = n(),
            freq_99=sum(heatwave_99)/n(),
            freq_95=sum(heatwave_95)/n(),
            freq_90=sum(heatwave_90)/n(),
            .groups="drop")%>%
  pivot_longer(cols = c(freq_99, freq_95, freq_90), values_to="heatwave_freq", names_to="quantile", names_prefix="freq_")%>%
  mutate(quantile=as.integer(quantile))

heatwave_sum_year_month<-temp_appt%>%
  group_by(major_city, Year, Month)%>%
  summarise(freq_99=sum(heatwave_99)/n(),
            freq_95=sum(heatwave_95)/n(),
            freq_90=sum(heatwave_90)/n(),
            .groups="drop")%>%
  pivot_longer(cols = c(freq_99, freq_95, freq_90), values_to="heatwave_freq", names_to="quantile", names_prefix="freq_")%>%
  mutate(quantile=as.integer(quantile))

cutoffs<-temp_appt%>%
  distinct(major_city, Month, temp_appt_q99, temp_appt_q95, temp_appt_q90)%>%
  pivot_longer(cols = c(temp_appt_q99, temp_appt_q95, temp_appt_q90), values_to="cutoff", names_to="quantile", names_prefix="temp_appt_q")%>%
  mutate(quantile=as.integer(quantile))

```

# Model changes over years
```{r}
model_dat<-heatwave_sum_year%>%
  filter(quantile==95)%>%
  mutate(Year_s=(Year-mean(Year))/sd(Year))

mf1<-lm(heatwave_freq ~ Year_s*major_city, data=model_dat)

summary(mf1)
```

# Heatwave data

```{r}
datatable(mutate(heatwave_dur, across(c(temp_appt_mean, intensity_mean), ~round(.x, 2))), rownames = FALSE, 
          options = list(lengthMenu = list(c(10,25,50,-1), c(10,25,50,"All"))))
```

# Plots 

Raw temperature vs apparent temperature
```{r}
ggplot(temp_appt, aes(x=temp_max, y=temp_appt))+
  geom_point(alpha=0.1)+
  geom_abline(slope=1, intercept=0, color="red", linetype=2)+
  facet_wrap(~Month)+
  theme_bw()
```

humidity by month
```{r}
ggplot(temp_appt, aes(x=Month, y=rh))+
  geom_boxplot()+
  facet_wrap(~major_city)+
  theme_bw()+
  theme(axis.text.x=element_text(angle=45, hjust=1))
```

dew point by month
```{r}
ggplot(temp_appt, aes(x=Month, y=dewpt))+
  geom_boxplot()+
  facet_wrap(~major_city)+
  theme_bw()+
  theme(axis.text.x=element_text(angle=45, hjust=1))
```

actvap point by month
```{r}
ggplot(temp_appt, aes(x=Month, y=actvap))+
  geom_boxplot()+
  facet_wrap(~major_city)+
  theme_bw()+
  theme(axis.text.x=element_text(angle=45, hjust=1))
```

## Apparent temperature data

```{r, fig.height=12, fig.width=10}
ggplot(temp_appt, aes(x=yday(date), y=temp_appt))+
  geom_point()+
  facet_grid(Year~major_city)+
  theme_bw()
```

## Heatwaves

### Plot apparent temperature cutoffs

```{r}
ggplot(cutoffs, aes(x=Month, y=cutoff, color=quantile))+
  geom_point()+
  facet_wrap(~major_city)+
  scale_color_viridis_c(breaks=c(90, 95, 99))+
  ylab("Apparent temperature cutoff (°C)")+
  theme_bw()+
  theme(axis.text.x=element_text(angle=45, hjust=1))
```

### Create table of temperature cutoffs

```{r}
datatable(mutate(cutoffs, cutoff=round(cutoff, 1)), rownames = FALSE, 
          options = list(lengthMenu = list(c(10,25,50,-1), c(10,25,50,"All"))))
```

### Plot temps with heat waves

```{r, fig.height=12, fig.width=10}
ggplot(temp_appt, aes(x=date, y=temp_appt))+
  geom_point()+
  geom_vline(data=filter(temp_appt, heatwave_90), aes(xintercept=date), color="chartreuse3")+
  geom_vline(data=filter(temp_appt, heatwave_95), aes(xintercept=date), color="darkorchid4")+
  geom_vline(data=filter(temp_appt, heatwave_99), aes(xintercept=date), color="red")+
  facet_grid(Month~major_city)+
  theme_bw()+
  theme(axis.text.x=element_text(angle=45, hjust=1))
```

### Plot heat wave frequency by month

```{r, fig.height=9, fig.width=10}
ggplot(heatwave_sum_month, aes(x=Month, y=heatwave_freq, fill=major_city))+
  geom_bar(position=position_dodge(), stat = "identity")+
  facet_wrap(~quantile)+
  scale_y_continuous(expand=expansion(0,0))+
  scale_fill_viridis_d()+
  ylab("Heatwave frequency")+
  theme_bw()+
  theme(axis.text.x=element_text(angle=45, hjust=1), legend.position = "bottom")
```

### Plot heat wave frequency by year

```{r, fig.height=9, fig.width=10}
ggplot(heatwave_sum_year, aes(x=Year, y=heatwave_freq, fill=major_city))+
  geom_bar(position=position_dodge(), stat = "identity")+
  facet_wrap(~quantile)+
  scale_y_continuous(expand=expansion(0,0))+
  scale_fill_viridis_d()+
  ylab("Heatwave frequency")+
  theme_bw()+
  theme(axis.text.x=element_text(angle=45, hjust=1), legend.position = "bottom")
```

### Plot heat wave frequency by year and month

```{r, fig.height=9, fig.width=10}
ggplot(heatwave_sum_year_month, aes(x=Year, y=heatwave_freq, fill=major_city))+
  geom_bar(position=position_dodge(), stat = "identity")+
  facet_grid(Month~quantile)+
  scale_y_continuous(expand=expansion(0,0))+
  scale_fill_viridis_d()+
  ylab("Heatwave frequency")+
  theme_bw()+
  theme(axis.text.x=element_text(angle=45, hjust=1), legend.position = "bottom")
```

### Heatwave duration

```{r}
ggplot(heatwave_dur, aes(x=Year, y=duration, color=major_city, group=interaction(Year, major_city)))+
  geom_jitter(height=0, alpha=0.3)+
  scale_color_viridis_d(name="City")+
  facet_wrap(~major_city)+
  ylab("Heatwave duration (days)")+
  theme_bw()
```

Aggregated by year
```{r}
ggplot(heatwave_dur_year, aes(x=Year, y=duration_mean, ymin=duration_mean-duration_sd, 
                              ymax=duration_mean+duration_sd, color=major_city, 
                              group=interaction(Year, major_city)))+
  geom_pointrange()+
  scale_color_viridis_d(name="City")+
  facet_wrap(~major_city)+
  ylab("Heatwave duration (days)")+
  theme_bw()
```
### Heatwave intensity

By Month
```{r}
ggplot(filter(temp_appt, heatwave_95), 
       aes(x=Month, y=heatwave_95_intensity, fill=major_city, group=interaction(Month, major_city)))+
  geom_boxplot()+
  scale_fill_viridis_d(name="City")+
  theme_bw()
```

By Year
```{r}
ggplot(filter(temp_appt, heatwave_95), 
       aes(x=Year, y=heatwave_95_intensity, fill=major_city, group=interaction(Year, major_city)))+
  geom_boxplot()+
  scale_fill_viridis_d(name="City")+
  theme_bw()
```

### Final plots

Temp cutoffs

```{r}
p<-ggplot(filter(cutoffs, quantile==95), aes(x=Month, y=cutoff, color=major_city))+
  geom_point(position=position_dodge(width=0.4), alpha=0.8)+
  scale_color_viridis_d(name="")+
  ylab("Apparent temperature cutoff (°C)")+
  theme_bw()+
  theme(axis.text.x=element_text(angle=45, hjust=1), legend.position = "bottom", plot.margin = margin(5, 40, 5, 5))
p

ggsave(plot=p, file="figs/air apparent temperature cutoffs.png", height=5, width=5)
```

Results
```{r, fig.height=9, fig.width=10}
p_y<-ggplot(filter(heatwave_sum_month, quantile==95), aes(x=Month, y=heatwave_freq, fill=major_city))+
  geom_bar(position=position_dodge(), stat = "identity")+
  scale_y_continuous(expand=expansion(0,0))+
  scale_fill_viridis_d(name="")+
  ylab("Heatwave frequency")+
  theme_bw()+
  theme(axis.text.x=element_text(angle=45, hjust=1), legend.position = "bottom")

p_m<-ggplot(filter(heatwave_sum_year, quantile==95), aes(x=Year, y=heatwave_freq, fill=major_city, color=major_city))+
    geom_bar(position=position_dodge(), stat = "identity")+
    geom_smooth(method="lm", se=F, show.legend = FALSE)+
    scale_y_continuous(expand=expansion(0,0))+
    scale_fill_viridis_d(name="", aesthetics=c("color", "fill"))+
    ylab("Heatwave frequency")+
    theme_bw()+
    theme(axis.text.x=element_text(angle=45, hjust=1), legend.position = "bottom")

p<-p_m/p_y+plot_layout(guides="collect") &
  theme(legend.position='bottom')
p

ggsave(plot=p, file="figs/air heatwave frequency.png", height=6, width=10)
```

### Write data to plot with water temp data
```{r}
city_boundaries<-zctas(starts_with = zips$zipcode)%>%
  left_join(zips, by=c("ZCTA5CE20"="zipcode"))%>%
  group_by(major_city)%>%
  summarise(population=sum(population))%>%
  st_transform(crs=26910)
  
st_write(city_boundaries, "shapefiles/SBDS city boundaries/SBDS city boundaries.shp", delete_dsn=TRUE)
# save(cutoffs, temp_appt, heatwave_sum_month, heatwave_sum_year, heatwave_dur, heatwave_dur_year, temp_appt, file = here::here("data_clean/air_temp_data_for_plots.Rdata"))
```
