---
title: "heatwave_analysis"
author: "Catarina Pien"
date: '2023-02-04'
output: html_document
editor_options: 
  chunk_output_type: console
---

# Heatwave Analysis for Water Temperature Data 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r}
library(dplyr)
library(lubridate)
library(tidyr)
library(ggplot2)
library(sf)
library(mapview)
```

## Read data - filter to stations of interest
```{r}
stations <- c("RRI", "CLL", "SRH", "GYS", "MRZ", "VER", "RIV", "JER")
tempdata <- readRDS("data_clean/Temp_filtered.rds") %>%
  filter(Station %in% stations) %>%
  rename(date = Date)
latlons <- read.csv("https://portal.edirepository.org/nis/dataviewer?packageid=edi.591.2&entityid=a059f5eea4f8500fe1a43566451ec593")
```

## Make map
```{r}
sta_locations_sf <- latlons %>%
  filter(Station %in% stations) %>%
  st_as_sf(coords = c("Longitude", "Latitude"),
           crs = 4326)
  
mapview::mapView(sta_locations_sf, zcol = "Station")
```

## Add parameters
```{r}
dailytemp <- tempdata %>%
  group_by(Station, date) %>%
  summarize(temp_min = min(Temp),
            temp_mean = mean(Temp),
            temp_max = max(Temp)) %>%
  mutate(Month = month(date),
         fMonth = factor(Month),
         Year = year(date),
         WY = ifelse(Month > 9, Year + 1, Year),
         Season = case_when(Month %in% c(10, 11, 12) ~ "Fall",
                            Month %in% c(1, 2,3) ~ "Winter",
                            Month %in% c(4, 5, 6) ~ "Spring",
                            Month %in% c(7, 8, 9) ~ "Summer")) %>%
  ungroup() 
```

## Look at data
```{r}
dailytemp %>%
  group_by(Station, Year) %>%
  summarize(n = n()) %>%
  ggplot() + geom_tile(aes(Year, Station, fill = n), color = "black") +
  theme_bw()
```

## Filter data

* All months 
* 2002 +
* Calculate quantiles for mean, max, min
* Classify as heatwave if 3+ days over quantiles
```{r}
temp_quantiles = dailytemp %>%
  filter(Year>2001) %>%
  #filter(Month %in% c(5, 6, 7, 8, 9, 10)) %>%
  dplyr::group_by(Station, Month, fMonth) %>%
  mutate(temp_max_q99 = quantile(temp_max, 0.99),
         temp_max_q95 = quantile(temp_max, 0.95),
         temp_max_q90 = quantile(temp_max, 0.90),
         temp_mean_q99 = quantile(temp_mean, 0.99),
         temp_mean_q95 = quantile(temp_mean, 0.95),
         temp_mean_q90 = quantile(temp_mean, 0.90),
         temp_min_q99 = quantile(temp_min, 0.99),
         temp_min_q95 = quantile(temp_min, 0.95),
         temp_min_q90 = quantile(temp_min, 0.90)) %>%
  ungroup() %>%
  dplyr::group_by(Station) %>%
  mutate(temp_max_l1 = lag(temp_max, 1, order_by = date),
         temp_max_l2 = lag(temp_max, 2, order_by = date),
         heatwave_99=if_else(temp_max>temp_max_q99 & temp_max_l1>temp_max_q99 & temp_max_l2>temp_max_q99, 1L, 0L),
         heatwave_95=if_else(temp_max>temp_max_q95 & temp_max_l1>temp_max_q95 & temp_max_l2>temp_max_q95, 1L, 0L),
         heatwave_90=if_else(temp_max>temp_max_q90 & temp_max_l1>temp_max_q90 & temp_max_l2>temp_max_q90, 1L, 0L),
         # min
         temp_min_l1 = lag(temp_min, 1, order_by = date),
         temp_min_l2 = lag(temp_min, 2, order_by = date),
         heatwave_99=if_else(temp_min>temp_min_q99 & temp_min_l1>temp_min_q99 & temp_min_l2>temp_min_q99, 1L, 0L),
         heatwave_95=if_else(temp_min>temp_min_q95 & temp_min_l1>temp_min_q95 & temp_min_l2>temp_min_q95, 1L, 0L),
         heatwave_90=if_else(temp_min>temp_min_q90 & temp_min_l1>temp_min_q90 & temp_min_l2>temp_min_q90, 1L, 0L),
         # mean
         temp_mean_l1 = lag(temp_mean, 1, order_by = date),
         temp_mean_l2 = lag(temp_mean, 2, order_by = date),
         heatwave_99=if_else(temp_mean>temp_mean_q99 & temp_mean_l1>temp_mean_q99 & temp_mean_l2>temp_mean_q99, 1L, 0L),
         heatwave_95=if_else(temp_mean>temp_mean_q95 & temp_mean_l1>temp_mean_q95 & temp_mean_l2>temp_mean_q95, 1L, 0L),
         heatwave_90=if_else(temp_mean>temp_mean_q90 & temp_mean_l1>temp_mean_q90 & temp_mean_l2>temp_mean_q90, 1L, 0L))
```

## Calculate percentiles

* Calculate heatwave frequency
```{r}
heatwave_sum_max<-temp_quantiles%>%
  group_by(Station, Month, fMonth)%>%
  summarise(freq_99=sum(heatwave_99)/n(),
            freq_95=sum(heatwave_95)/n(),
            freq_90=sum(heatwave_90)/n(),
            .groups="drop")%>%
  pivot_longer(cols = c(freq_99, freq_95, freq_90), values_to="heatwave_freq", names_to="quantile", names_prefix="freq_")%>%
  mutate(quantile=as.integer(quantile),
         Month=month(Month, label=T),
         fMonth = factor(Month))


heatwave_sum_max_month_year<-temp_quantiles%>%
  group_by(Station, Year, Month, fMonth)%>%
  summarise(freq_99=sum(heatwave_99)/n(),
            freq_95=sum(heatwave_95)/n(),
            freq_90=sum(heatwave_90)/n(),
            .groups="drop")%>%
  pivot_longer(cols = c(freq_99, freq_95, freq_90), values_to="heatwave_freq", names_to="quantile", names_prefix="freq_")%>%
  mutate(quantile=as.integer(quantile),
         Month=month(Month, label=T),
         fMonth = factor(Month))

heatwave_sum_max_year<-temp_quantiles%>%
  group_by(Station, Year)%>%
  summarise(freq_99=sum(heatwave_99)/n(),
            freq_95=sum(heatwave_95)/n(),
            freq_90=sum(heatwave_90)/n(),
            .groups="drop")%>%
  pivot_longer(cols = c(freq_99, freq_95, freq_90), values_to="heatwave_freq", names_to="quantile", names_prefix="freq_")%>%
  mutate(quantile=as.integer(quantile))
```

## Plots

* Organize station west to east with color gradient
* Include a map of air and water temp overlaid
* Double check my year code for percentages


### Plot quantile values
```{r, fig.height = 6, fig.width = 8}
temp_quantile_values = dailytemp %>%
  filter(Year>2001) %>%
  #filter(Month %in% c(5, 6, 7, 8, 9, 10)) %>%
  dplyr::group_by(Station, Month) %>%
  summarise(temp_max_q99 = quantile(temp_max, 0.99),
         temp_max_q95 = quantile(temp_max, 0.95),
         temp_max_q90 = quantile(temp_max, 0.90)) %>%
  pivot_longer(cols = c(temp_max_q99, temp_max_q95, temp_max_q90), values_to = "temp", names_to = "quantile", names_prefix = "temp_max_") %>%
  mutate(fMonth = factor(Month))

ggplot(temp_quantile_values) + 
  geom_point(aes(x = fMonth, y = temp, color = quantile)) +
  facet_wrap(~Station) + 
  viridis::scale_color_viridis(option = "turbo", discrete = TRUE) + 
  theme_bw() 
```

### Plot temperature values with quantiles
```{r, fig.height=8, fig.width=10}
# with quantile values
ggplot() + 
  geom_jitter(data = temp_quantiles, aes(x = fMonth, y = temp_max), color = "gray80", shape = 2, size = 0.5) +
  #geom_line(data = temp_quantile_values, aes(x = fMonth, y = temp, color = quantile))+
  geom_point(data = temp_quantile_values, aes(x = fMonth, y = temp, color = quantile), size = 2) +
  facet_grid(Year~Station) + 
  viridis::scale_color_viridis(option = "turbo", discrete = TRUE) + 
  theme_bw() +
  theme(legend.position = "top")


ggplot(temp_quantiles, aes(x=yday(date), y=temp_max))+
  geom_point()+
  facet_grid(Year~Station)+
  theme_bw()
```

### Plot temps with shaded periods of heat waves
```{r, fig.height=8, fig.width=10}
ggplot(temp_quantiles, aes(x=date, y=temp_max))+
  geom_point()+
  geom_vline(data=filter(temp_quantiles, heatwave_90==1L), aes(xintercept=date), color="chartreuse3")+
  geom_vline(data=filter(temp_quantiles, heatwave_95==1L), aes(xintercept=date), color="darkorchid4")+
  geom_vline(data=filter(temp_quantiles, heatwave_99==1L), aes(xintercept=date), color="red")+
  facet_grid(Station~Month)+
  theme_bw()+
  theme(axis.text.x=element_text(angle=45, hjust=1))
```

### Plot heat wave frequency by month
```{r, fig.height=6, fig.width=10}
ggplot(heatwave_sum_max, aes(x=Month, y=heatwave_freq, fill=Station))+
  geom_bar(position=position_dodge(), stat = "identity")+
  facet_wrap(~quantile)+
  scale_y_continuous(expand=expansion(0,0))+
  scale_fill_viridis_d()+
  theme_bw()+
  theme(axis.text.x=element_text(angle=45, hjust=1))
```

### Plot heat wave frequency by year
```{r, fig.height = 6, fig.width = 10}
ggplot(heatwave_sum_max_year, aes(x=factor(Year), y=heatwave_freq, fill=Station))+
  geom_bar(position=position_dodge(), stat = "identity")+
  facet_wrap(~quantile)+
  scale_y_continuous(expand=expansion(0,0))+
  scale_fill_viridis_d()+
  theme_bw()+
  theme(axis.text.x=element_text(angle=45, hjust=1))
```

### Plot heat wave frequency by year, month, station, and quantile type
```{r, fig.height = 8, fig.width = 10}
ggplot(heatwave_sum_max_month_year, aes(x=factor(Year), y=heatwave_freq, fill=Station))+
  geom_bar(position=position_dodge(), stat = "identity")+
  facet_grid(fMonth~quantile)+
  scale_y_continuous(expand=expansion(0,0))+
  scale_fill_viridis_d()+
  theme_bw()+
  theme(axis.text.x=element_text(angle=45, hjust=1))
```

### Plot heat wave frequency by space
```{r}
heatwaves_sf <- right_join(sta_locations_sf, heatwave_sum_max_year ) %>%
  filter(quantile == "95", fMonth == "Jul")

ggplot() + 
  geom_sf(data = heatwaves_sf, aes(color = heatwave_freq)) +
  facet_wrap(~Year)+
  theme(axis.text.x=element_text(angle=45, hjust=1))
```

