---
title: "heatwave_analysis"
author: "Catarina Pien"
date: '2023-02-04'
output: html_document
editor_options: 
  chunk_output_type: console
---

# Heatwave Analysis for Water Temperature Data 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r}
library(dplyr)
library(lubridate)
library(tidyr)
library(ggplot2)
library(sf)
library(mapview)
library(DT)
library(patchwork)
library(viridis)
library(slider)
library(imputeTS)
library(purrr)
library(mgcv)
source("heatwave functions.R")
```

## Read data - filter to stations of interest
```{r}
stations <- c("RRI", "CLL", "SRH", "GYS", "MRZ", "VER", "RIV", "JER")
tempdata <- readRDS("data_clean/Temp_filtered.rds") %>%
  filter(Station %in% stations) %>%
  rename(date = Date)
latlons <- read.csv("https://portal.edirepository.org/nis/dataviewer?packageid=edi.591.2&entityid=a059f5eea4f8500fe1a43566451ec593")
```

## Make map
```{r}
 sta_locations_sf <- latlons %>%
   filter(Station %in% stations) %>%
   st_as_sf(coords = c("Longitude", "Latitude"),
           crs = 4326)
 
 mapview::mapView(sta_locations_sf, zcol = "Station")
 
station_order <- c("MRZ", "GYS", "CLL", "JER", "RIV", "SRH", "RRI", "VER")
```

## Add parameters
```{r}
dailytemp <- tempdata %>%
  filter(year(date)>2000)%>%
  group_by(Station, date) %>%
  summarize(temp_min = min(Temp),
            temp_mean = mean(Temp),
            temp_max = max(Temp), .groups="drop") %>%
  complete(Station, date) 

dailytemp_wide<-dailytemp%>%
  select(-temp_min, -temp_mean)%>%
  pivot_wider(names_from=Station, values_from=temp_max)%>%
  mutate(year=year(date),
         doy=yday(date))
```

## imputation
```{r fig.width=12}
dailytemp_imp<-dailytemp%>%
  group_by(Station)%>%
  arrange(date)%>%
  mutate(temp_min = na_ma(temp_min, k = 7, weighting = "exponential", maxgap = 3),
         temp_mean = na_ma(temp_mean, k = 7, weighting = "exponential", maxgap = 3),
         temp_max = na_ma(temp_max, k = 7, weighting = "exponential", maxgap = 3))%>%
  ungroup()%>%
  mutate(Month = month(date),
         fMonth = month(date, label = TRUE),
         fMonth = factor(fMonth),
         Year = year(date),
         WY = ifelse(Month > 9, Year + 1, Year),
         doy = yday(date),
         Season = case_when(Month %in% c(10, 11, 12) ~ "Fall",
                            Month %in% c(1, 2,3) ~ "Winter",
                            Month %in% c(4, 5, 6) ~ "Spring",
                            Month %in% c(7, 8, 9) ~ "Summer"))
```

## plot imputation
```{r fig.width=12}
imp_plot<-function(station){
p<-ggplot_na_imputations(filter(dailytemp, Station==station)$temp_max, filter(dailytemp_imp, Station==station)$temp_max,  x_axis_labels=filter(dailytemp_imp, Station==station)$date)+ggtitle(station)
print(p)
}

walk(stations, imp_plot)
```


## Look at data
```{r}
dailytemp_imp %>%
  group_by(Station, Year) %>%
  summarize(n = n()) %>%
  ggplot() + geom_tile(aes(Year, Station, fill = n), color = "black") +
  theme_bw()
```

## Filter data

* All months 
* 2002 +
* Calculate quantiles for mean, max, min
* Classify as heatwave if 3+ days over quantiles
```{r}
dailytemp_2001<-dailytemp_imp%>%
  filter(Year>2001 & !is.na(temp_max))

temp_quantiles = dailytemp_2001  %>%
  rowwise()%>%
  mutate(data_doy=list(dailytemp_2001$temp_max[dailytemp_2001$doy%in%date_slider(date) & dailytemp_2001$Station==Station]),
         temp_max_q99=quantile(data_doy, 0.99),
         temp_max_q95=quantile(data_doy, 0.95),
         temp_max_q90=quantile(data_doy, 0.90))%>%
  select(-data_doy)%>%
  ungroup()%>%
  dplyr::group_by(Station) %>%
  mutate(heatwave_99=heatwave_detector(temp_max, temp_max_q99),
         heatwave_95=heatwave_detector(temp_max, temp_max_q95),
         heatwave_90=heatwave_detector(temp_max, temp_max_q90),
         heatwave_95_intensity=if_else(heatwave_95, temp_max/temp_max_q95, NA),
         heatwave_90_intensity=if_else(heatwave_90, temp_max/temp_max_q90, NA)) %>%
  ungroup()%>%
  arrange(Station,date)

temp_quantiles$Station <- factor(temp_quantiles$Station, levels = station_order)
```

## Heatwave duration
```{r}
# sam's code ----------------------
heatwave_dur <- temp_quantiles %>%
  filter(heatwave_90) %>%
  group_by(Station) %>%
  mutate(Lag = date-lag(date, order_by = date)) %>%
  ungroup() %>%
  mutate(Start= if_else(is.na(Lag) | Lag > 1, TRUE, FALSE),
         Heatwave_id = 1:n(),
         Heatwave_id = if_else(Start, Heatwave_id, NA_integer_),
         Heatwave_id=as.integer(as.factor(Heatwave_id)))%>%
  fill(Heatwave_id, .direction="down")%>%
  group_by(Station, Heatwave_id)%>%
  summarise(duration=n(),
            Date=min(date),
            temp_mean=mean(temp_max),
            intensity_mean=mean(heatwave_90_intensity),
            .groups="drop")%>%
  mutate(Month=month(Date, label=T),
         Year=year(Date))

heatwave_dur_year<-heatwave_dur%>%
  group_by(Year, Station)%>%
  summarise(duration_mean=mean(duration),
            duration_sd=sd(duration),
            .groups="drop")%>%
  mutate(duration_sd=if_else(is.na(duration_sd), 0, duration_sd))
```

## Calculate percentiles

* Calculate heatwave frequency
```{r}
heatwave_sum_month<-temp_quantiles%>%
  group_by(Station, Month, fMonth)%>%
  summarise(n = n(),
            freq_99=sum(heatwave_99)/n(),
            freq_95=sum(heatwave_95)/n(),
            freq_90=sum(heatwave_90)/n(),
            .groups="drop")%>%
  tidyr::pivot_longer(cols = c(freq_99, freq_95, freq_90), values_to="heatwave_freq", names_to="quantile", names_prefix="freq_")%>%
  mutate(quantile=as.integer(quantile),
         Month=lubridate::month(Month, label=T),
         fMonth = factor(Month)) %>%
  filter(quantile == 90)


heatwave_sum_max_month_year<-temp_quantiles%>%
  group_by(Station, Year, Month, fMonth)%>%
  summarise(freq_99=sum(heatwave_99)/n(),
            freq_95=sum(heatwave_95)/n(),
            freq_90=sum(heatwave_90)/n(),
            .groups="drop")%>%
  pivot_longer(cols = c(freq_99, freq_95, freq_90), values_to="heatwave_freq", names_to="quantile", names_prefix="freq_")%>%
  mutate(quantile=as.integer(quantile),
         Month=month(Month, label=T),
         fMonth = factor(Month)) %>%
  filter(quantile == 90)

heatwave_sum_max_year<-temp_quantiles%>%
  group_by(Station, Year)%>%
  summarise(n = n(),
            freq_99=sum(heatwave_99)/n(),
            freq_95=sum(heatwave_95)/n(),
            freq_90=sum(heatwave_90)/n(),
            .groups="drop")%>%
  pivot_longer(cols = c(freq_99, freq_95, freq_90), values_to="heatwave_freq", names_to="quantile", names_prefix="freq_")%>%
  mutate(quantile=as.integer(quantile)) %>%
  filter(quantile == 90)

cutoffs<-temp_quantiles%>%
  distinct(Station, fMonth, doy, temp_max_q99, temp_max_q95, temp_max_q90)%>%
  pivot_longer(cols = c(temp_max_q99, temp_max_q95, temp_max_q90), values_to="cutoff", names_to="quantile", names_prefix="temp_max_q")%>%
  mutate(quantile=as.integer(quantile))
```

## Model changes over years
```{r}
w_model_dat<-heatwave_sum_max_year%>%
  filter(quantile==90)%>%
  mutate(Year_s=(Year-mean(Year))/sd(Year))

w_model_dat2<-heatwave_sum_max_month_year%>%
  filter(quantile==90)%>%
  mutate(Year_s=(Year-mean(Year))/sd(Year))

mf1<-lm(heatwave_freq ~ Year_s + factor(Station) , data=w_model_dat)
summary(mf1)

mf2<-lm(heatwave_freq ~ Year_s* factor(Station), data=w_model_dat)
summary(mf2)

anova(mf1)
par(mfrow = c(2,2))
plot(mf1)

library(lmerTest)
library(lme4)
mf3 <- lmer(heatwave_freq ~ Year_s * factor(Station) + (1|fMonth), data=w_model_dat2)
summary(mf3)
anova(mf3, type = 2)

plot(mf3)


summary(mf1)
```

### Plot temperature values
```{r, fig.height=8, fig.width=10}
# with quantile values
# ggplot() + 
#   geom_jitter(data = temp_quantiles, aes(x = fMonth, y = temp_max), color = "gray80", shape = 2, size = 0.5) +
#   #geom_line(data = temp_quantile_values, aes(x = fMonth, y = temp, color = quantile))+
#   geom_point(data = temp_quantile_values, aes(x = fMonth, y = temp, color = quantile), size = 2) +
#   facet_grid(Year~Station) + 
#   viridis::scale_color_viridis(option = "turbo", discrete = TRUE) + 
#   theme_bw() +
#   theme(legend.position = "top")


ggplot(temp_quantiles, aes(x=yday(date), y=temp_max))+
  geom_point()+
  facet_grid(Year~Station)+
  theme_bw()
```

### Plot temps with shaded periods of heat waves
```{r, fig.height=8, fig.width=10}
ggplot(temp_quantiles, aes(x=date, y=temp_max))+
  geom_point()+
  # geom_vline(data=filter(temp_quantiles, heatwave_90==1L), aes(xintercept=date), color="chartreuse3")+
  geom_vline(data=filter(temp_quantiles, heatwave_90), aes(xintercept=date), color="darkorchid4")+
  # geom_vline(data=filter(temp_quantiles, heatwave_99==1L), aes(xintercept=date), color="red")+
  facet_grid(Station~Month)+
  theme_bw()+
  theme(axis.text.x=element_text(angle=45, hjust=1))
```

### *Plot heat wave frequency by month
```{r, fig.height=6, fig.width=10}
ggplot(heatwave_sum_max_month_year, aes(x=fMonth, y=heatwave_freq, fill=Station))+
  geom_bar(position=position_dodge(), stat = "identity", color = "black")+
  # facet_wrap(~quantile)+
  scale_y_continuous(expand=expansion(0,0))+
  scale_fill_viridis_d()+
  theme_bw()+
  theme(axis.text.x=element_text(angle=45, hjust=1))
```

### *Plot heat wave frequency by year
```{r, fig.height = 6, fig.width = 10}
ggplot(heatwave_sum_max_year, aes(x=factor(Year), y=heatwave_freq, fill=Station))+
  geom_bar(position=position_dodge(), stat = "identity", color = "black")+
  # facet_wrap(~quantile)+
  scale_y_continuous(expand=expansion(0,0))+
  scale_fill_viridis_d()+
  theme_bw()+
  theme(axis.text.x=element_text(angle=45, hjust=1))
```

### Plot heat wave frequency by year, month, station, and quantile type
```{r, fig.height = 8, fig.width = 10}
ggplot(heatwave_sum_max_month_year, aes(x=factor(Year), y=heatwave_freq, fill=Station))+
  geom_bar(position=position_dodge(), stat = "identity")+
  facet_grid(fMonth~.)+
  scale_y_continuous(expand=expansion(0,0))+
  scale_fill_viridis_d()+
  theme_bw()+
  theme(axis.text.x=element_text(angle=45, hjust=1), legend.position = "bottom", axis.title.x = element_blank(), plot.margin = margin(5, 40, 5, 5))
```

### Heatwave intensity
By Month
```{r}
temp_hw <- temp_quantiles %>% filter(heatwave_90)
(p_int_month <- ggplot(filter(temp_quantiles, heatwave_90), 
                       aes(x=fMonth, y=heatwave_90_intensity, fill=Station, group=interaction(fMonth, Station)))+
    geom_boxplot()+
    scale_fill_viridis(name="Station", option = "turbo", discrete = TRUE)+
    ylab("Heatwave Intensity")+
    theme_bw() +
    theme(axis.text.x=element_text(angle=45, hjust=1), legend.position = "bottom", axis.title.x = element_blank(), plot.margin = margin(5, 40, 5, 5)))
```

By Year
```{r}
(p_int_year <- ggplot(filter(temp_quantiles, heatwave_90), 
                      aes(x=Year, y=heatwave_90_intensity, fill=Station, group=interaction(Year, Station)))+
   geom_boxplot()+
   scale_fill_viridis_d(name="Station", option = "turbo")+
   scale_x_continuous(breaks = seq(2002, 2022, 1)) + 
   ylab("Heatwave Intensity")+
   theme_bw()+
   theme(axis.text.x=element_text(angle=45, hjust=1), legend.position = "bottom", axis.title.x = element_blank(), plot.margin = margin(5, 40, 5, 5)))
```

### *Duration
```{r}
# Comparing years - station facet
ggplot(heatwave_dur_year, aes(x=Year, y=duration_mean, ymin=duration_mean-duration_sd, 
                              ymax=duration_mean+duration_sd, color=Station, 
                              group=interaction(Year, Station)))+
  geom_pointrange()+
  scale_color_viridis_d(name="Station", option = "turbo")+
  facet_wrap(~Station)+
  ylab("Heatwave duration (days)")+
  theme_bw()
```

## More Finalized Plots (Final ones in combine_air_water_plots.Rmd)

```{r}
save(cutoffs, heatwave_sum_month, heatwave_sum_max_year, heatwave_dur, heatwave_dur_year, temp_quantiles, file = here::here("data_clean/water_temp_data_for_plots.Rdata"))
```

### Temp cutoffs

```{r}
p<-ggplot(filter(cutoffs, quantile==90), aes(x=doy, y=cutoff, color=Station, group=Station))+
  geom_line(alpha=0.8)+
  scale_color_viridis_d(name="")+
  ylab("Temperature cutoff (°C)")+
  xlab("Day of year")+
  theme_bw()+
  theme(axis.text.x=element_text(angle=45, hjust=1), legend.position = "bottom", plot.margin = margin(5, 40, 5, 5))
p

# ggsave(plot=p, file="figs/temperature cutoffs.png", height=5, width=5)
```


### Proportion
```{r, fig.height=9, fig.width=10}

p_m<-ggplot(filter(heatwave_sum_month, quantile==90), aes(x=Month, y=heatwave_freq, fill=Station))+
  geom_bar(position=position_dodge(), stat = "identity")+
  scale_y_continuous(expand=expansion(0,0))+
  scale_fill_viridis_d(name="", option = "turbo")+
  ylab("Heatwave frequency")+
  theme_bw()+
  theme(axis.text.x=element_text(angle=45, hjust=1), legend.position = "bottom", 
        axis.title.x = element_blank())

p_y<-ggplot(filter(heatwave_sum_max_year, quantile==90), aes(x=Year, y=heatwave_freq, fill=Station))+
  geom_bar(position=position_dodge(), stat = "identity")+
  scale_y_continuous(expand=expansion(0,0))+
  scale_x_continuous(breaks = seq(2002, 2022, 1)) + 
  scale_fill_viridis_d(name="", option = "turbo")+
  ylab("Heatwave frequency")+
  theme_bw()+
  theme(axis.text.x=element_text(angle=45, hjust=1), legend.position = "bottom",
        axis.title.x = element_blank())

p<-p_y/p_m+plot_layout(guides="collect") &
  theme(legend.position='bottom')
p

# png("figs/water heatwave results.png", height = 6, width = 10, units = "in", pointsize = 9,
# res=300)
# p
# dev.off()

```

### Intensity
```{r}
p_int<-p_int_year/p_int_month+plot_layout(guides="collect") &
  theme(legend.position='bottom')
p_int

# png("figs/water_intensity.png", height = 8, width = 10, units = "in", pointsize = 9,
# res=300)
# p_int
# dev.off()
```

### Duration

```{r}
ggplot(heatwave_dur_year, aes(x=Year, y=duration_mean, ymin=duration_mean-duration_sd, 
                              ymax=duration_mean+duration_sd, color=Station, 
                              group=interaction(Year, Station)))+
  geom_pointrange()+
  scale_color_viridis_d(name="Station", option = "turbo")+
  facet_wrap(~Station)+
  ylab("Heatwave duration (days)")+
  theme_bw()
```


## Summary tables
```{r, include = FALSE}
cutoffs_90 <- filter(cutoffs, quantile == 90)
cutoffs_summary <- cutoffs_90 %>%
  group_by(fMonth) %>%
  summarize(mean = mean(cutoff),
            se = sd(cutoff)/sqrt(n()))

heatwave <- temp_quantiles %>%
  filter(heatwave_90)

freq_month <- heatwave_sum_month %>%
  mutate(
    days = n*heatwave_freq,
    mean_overall = mean(heatwave_freq, na.rm = TRUE)) %>%
  group_by(Month, mean_overall) %>%
  reframe(mean = mean(heatwave_freq, na.rm = TRUE),
          se = sd(heatwave_freq, na.rm= TRUE)/sqrt(n()),
          mean_days = mean(days, na.rm = TRUE)) %>%
  distinct()

freq_year <- heatwave_sum_max_year %>%
  mutate(days = n*heatwave_freq,
         mean_overall = mean(heatwave_freq, na.rm = TRUE),
         days_overall = sum(n, na.rm = TRUE) * mean_overall,
         fYear = factor(Year),
         heatwave_freq = ifelse(is.na(heatwave_freq), 0, heatwave_freq)) %>%
  group_by(fYear, mean_overall, days_overall) %>%
  reframe(mean = mean(heatwave_freq, na.rm = TRUE),
          se = sd(heatwave_freq, na.rm= TRUE)/sqrt(n()),
          mean_days = mean(days, na.rm = TRUE)) %>%
  distinct()

int_month <- temp_quantiles %>%
  mutate(mean_overall = mean(heatwave_90_intensity, na.rm = TRUE),
         se_overall = sd(heatwave_90_intensity, na.rm = TRUE)/sqrt(n())) %>%
  group_by(fMonth, mean_overall, se_overall) %>%
  reframe(mean = mean(heatwave_90_intensity, na.rm = TRUE),
          se = sd(heatwave_90_intensity, na.rm = TRUE)/sqrt(n()))

int_year <- temp_quantiles %>%
  group_by(Year) %>%
  reframe(mean = mean(heatwave_90_intensity, na.rm = TRUE),
          se = sd(heatwave_90_intensity, na.rm = TRUE)/sqrt(n()),
          median = median(heatwave_90_intensity, na.rm = TRUE)) %>%
  arrange(mean)
```

