---
title: "visualize_data"
author: "Catarina Pien"
date: '2022-10-04'
output: html_document
editor_options: 
  chunk_output_type: console
---
# Heatwaves Analysis
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r,echo = FALSE, message = FALSE, warning = FALSE}
library(dplyr)
library(ggplot2)
library(lubridate)
library(tidyr)
```



## Questions
### Logistics 

1. What is the definition of a heatwave? 
    a) **“The World Meteorological Organization defines it as five or more consecutive days during which the daily maximum temperature surpasses the average maximum temperature by 5 °C (9 °F) or more.”**
    b) **this indicator defines a heat wave as a period of two or more consecutive days where the daily minimum apparent temperature at a particular weather station is higher than the 85th percentile of historical July and August temperatures for that city. Historical July and August baseline temperatures are analyzed for a base period of 1981–2010, which NCEI chose for consistency with other climatology metrics. Using the 85th percentile of July and August temperatures results in a threshold that equates to the nine hottest July and August days in an average year during that 30-year window. These are likely among the nine hottest days of the year.**
    c) What is the average maximum temperature? Delta-wide mean over 10 years? Just July and August?
      * 2+ days over 90th percentile mean/max/min monthly temp
    d) minimum or maximum temperatures? (try all)
2. What stations do we want to look at? 
  - longest running stations, 4 to 5 stations that represent across Delta
  - Use same regions/stations for air

3. Air and water temp 
  - need humidity data too 
  - scripps data for air (zip codes)

4. How many/what years?
  - last 10 to 20 years
### Analyses/Visualizations

4. When do heatwaves occur? (Seasonality)
5. Where do heatwaves occur? (Regions)
6. Has frequency, duration (hours vs total days in a row), length of heat wave season, intensity changed over time?
* trends over time 
* below freezing days (air temp) + consecutive below freezing days + intensity freezing days (mean temp) 

### References

* https://www.epa.gov/climate-indicators/climate-change-indicators-heat-waves
* Habeeb, D., J. Vargo, and B. Stone, Jr. 2015. Rising heat wave trends in large U.S. cities. Nat. Hazards 76(3):1651–1665. doi:10.1007/s11069-014-1563-z.

## Read data
```{r, echo = FALSE}
airtemp <- readRDS("data_raw/AirTemp.rds")
sta <- unique(airtemp$station)
hist(airtemp$airtemp_F)
glimpse(airtemp)

wtemp <- readRDS("data_raw/tempToUse_20210811.rds") %>%
  filter(Station %in% sta) %>%
  rename(station = Station,
         datetime = Datetime, 
         wtemp = Temp)
glimpse(wtemp)
unique(wtemp$station)

temps <- left_join(wtemp, airtemp, by = c("datetime", "station"))
subset <- filter(temps, WY >2018)
```

```{r}
ggplot(subset) + geom_point(aes(x = airtemp_C, y = wtemp)) + facet_wrap(~station)
```

## Calculate long term averages
```{r, echo = FALSE}
maxtemps <- temps %>%
  group_by(Region, WYType2_Sac, station, year, WY, month, date) %>%
  summarize(maxAir = max(airtemp_C),
            maxWater = max(wtemp)) %>%
  ungroup()

mintemps <- temps %>%
  group_by(Region, WYType2_Sac, station, year, WY, month, date) %>%
  summarize(minAir = min(airtemp_C),
            minWater = min(wtemp)) %>%
  ungroup()

# based on min
percentileTemps <- mintemps %>%
  filter(month %in% c(7, 8)) %>%
  group_by(station) %>%
  summarize(avgAir = quantile(minAir, 0.85),
           avgWater = quantile(minWater, 0.85)) %>%
  ungroup() %>%
  mutate(type = "percentile")

meanmaxannual <- maxtemps %>%
  group_by(station) %>%
  summarize(avgAir = mean(maxAir, na.rm = TRUE),
            avgWater = mean(maxWater, na.rm = TRUE)) %>%
  ungroup()%>%
  mutate(type = "annual")

meanmax6month <- maxtemps %>%
  filter(month %in% c(5, 6, 7, 8, 9, 10)) %>%
  group_by(station) %>%
  summarize(avgAir = mean(maxAir, na.rm = TRUE),
            avgWater = mean(maxWater, na.rm = TRUE)) %>%
  ungroup()%>%
  mutate(type = "sixmonth")

meanmax2month <- maxtemps %>%
  filter(month %in% c(7, 8)) %>%
  group_by(station) %>%
  summarize(avgAir = mean(maxAir, na.rm = TRUE),
            avgWater = mean(maxWater, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(type = "twomonth")

# Compare all the data
alldata <- bind_rows(meanmaxannual, meanmax6month, meanmax2month, percentileTemps) 
```

Compare different averages 
```{r, echo = TRUE}
plot_air <- ggplot(alldata) + geom_point(aes(station, avgAir, color = type), size = 2) + theme_bw()
plot_water <- ggplot(alldata) + geom_point(aes(station, avgWater, color = type), size = 2) + theme_bw()

library(patchwork)
plot_air + plot_water + plot_layout(guides = "collect")
```

